Bayesian Learning of Degenerate Linear~Gaussian~State~Space~Models using~Markov~Chain~Monte~Carlo

### 2015-09-18 notes/paper-degenerate-linear-systems.tex
About halfwy through rewrite

The Gibbs sampler approach developed in \cite{Wills2012} is effective on the basic flavour of linear Gaussian state space models. However, it is not able to handle the case in which the transition model is degenerate, i.e. when the transition covariance matrix is not full rank. In this paper we extend the MCMC learning framework to encompass this scenario. This is achieved by appropriate factorisations of the covariance matrix which allow us still to exploit the fast-mixing Gibbs kernel. Furthermore, an extended distribution construction is introduced which enables the sampler to learn the rank of the transition covariance matrix.

@@

\subsection{MCMC for Basic Linear Gaussian Models}

@@

\section{Degenerate Transitions} \label{sec:degenerate_transition_models}

A commonly encountered variation on the basic linear Gaussian state space model is as follows,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \lgdm \tn{\ti} \label{eq:degenerate_transition} \\
 \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\idmat}      .
\end{IEEEeqnarray}
%
This is almost equivalent, with $\lgtv=\lgdm\lgdm\tr$. However, whereas before $\lgtv$ was required to be positive definite, this is not necessarily the case for $\lgdm\lgdm\tr$. When this occurs, the conventional transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgdm)$ is not defined, and the sampling operations underlying our Gibbs sampler are no longer valid. Degenerate models such as these arise in tracking applications (see \cite{Maskell2004} for a discussion of this phenomenon); when non-Markovian models, such as autoregressive process, are converted to state space form; and when it is natural to parameterise a system with more latent states than there are degrees of freedom (see section~\ref{sec:mocap}). In this section, we introduce a parameterisation for such degenerate transition models which can uniquely represent any such system, and show how MCMC moves may be implemented for learning.

Attempting to learn $\lgdm$ directly is problematic, as this matrix is not identifiable. The distribution of $\lgdm\tn{\ti}$ is identical to that of $\lgdm \Xi \tn{\ti}$ where $\Xi$ is an arbitrary orthogonal matrix. Therefore, there is not a unique value of $\lgdm$ which maximises the probability of the state sequence $\den(\ls{1:\timax}|\lgtm,\lgdm)$. Instead, as in the basic model, we parameterise the system with a $\lsd\times\lsd$ precision matrix $\lgtp$ which is positive semi-definite with rank $\tnd$. For a particular value of $\lgtp$, $\lgdm$ may be any matrix such that,
%
\begin{IEEEeqnarray}{rCl}
 \lgdm\lgdm\tr & = & \lgtp\pinv     ,
\end{IEEEeqnarray}
%
where $\pinv$ denote the pseudo-inverse.

Learning degenerate transition models is challenging because each transition, $(\ls{\ti}-\lgtm\ls{\ti-1})$, is constrained to lie in a particular $\tnd$-dimensional subspace within the state space; specifically, within the column space of $\lgdm$. This subspace is fixed conditional on any sampled state trajectory $\ls{1:\timax}$ (provided that $\timax>\tnd$). This makes standard Gibbs sampling from $\postden(\lgtp|\lgtm,\ls{1:\timax})$ and $\postden(\lgtm|\lgtp,\ls{1:\timax})$ insufficient, since there are values of $\lgtm$ and $\lgtp$ which can never be reached.

In order to allow the Markov chain to fully explore the parameter space, we introduce Markov kernels which target alternative conditional distributions, specifically $\postden(\lgtm, \ls{1:\timax}|\lgtp)$ and $\postden(\lgtp|\lgtm)$. Drawing from these distributions exactly is not possible, but we can use Metropolis-Hastings. Naively, we might simply try to propose changes to the elements of $\lgtm$ and $\lgtp$, but this is very inefficient. Better schemes can be formulated by exploiting the following factorisations.

For a rank-$\tnd$ positive semi-definite precision matrix $\lgtp$,
%
\begin{IEEEeqnarray}{rCl}
 \lgtp & = & \tnmo \tnmd \tnmo\tr \nonumber      ,
\end{IEEEeqnarray}
%
where $\tnmd = \diag([\tnmev{1},\tnmev{2},\dots])$ is the diagonal matrix of $\tnd$ \emph{positive} eigenvalues (such that $\tnmev{1} > \tnmev{2} > \dots > \tnmev{\tnd}$), and $\tnmo$ is a $\lsd\times\tnd$ matrix of orthonormal eigenvectors. This is a standard eigendecomposition. As in \cite{Muirhead1982}, the first non-zero element in each column of $\tnmo$ is required to be positive for each $i$. This resolves the sign ambiguity in the eigenvectors and thus ensures that the transformation from $\lgtp$ to $\{\tnmo,\tnmd\}$ is bijective.

An alternative decomposition will also be useful. For a rank-$\tnd$ positive semi-definite precision matrix $\lgtp$,
%
\begin{IEEEeqnarray}{rCl}
 \lgtp & = & \tnmocrossred \tnmf \tnmocrossred\tr \label{eq:degenerate_covariance_factorisation}      ,
\end{IEEEeqnarray}
%
where,
%
$\tnmf$ is a $\tnd\times\tnd$ positive definite matrix and $\tnmocrossred$ is a $\lsd\times\tnd$ matrix of orthonormal vectors subject to certain conditions. The transformation from $\lgtp$ to $\{\tnmocrossred,\tnmf\}$ is also bijective. See appendix~\ref{app:givens-factorisation} for details.

@@

In the basic linear model, when $\lgtp$ was positive definite, we used a Wishart prior for this matrix. The singular Wishart distribution is an extension over the space of positive semi-definite matrices \cite{Uhlig1994}, and has the following density with respect to the Lebesgue measure on the distinct matrix elements, conditional on a known value of $\tnd$,
%
\begin{IEEEeqnarray}{rCl}
 \den(\lgtp|\tnd) & = & \wishartden{\lgtp}{\tnd}{\Ppscale} \nonumber \\
 & \propto & \determ{\tnmd}^{\frac{\tnd-\lsd-1}{2}} \exp\left( -\half \trace\left[ \Ppscale\inv \lgtp \right] \right)     .
\end{IEEEeqnarray}
%
The prior is completed by specifying a distribution on the integer $\tnd \in [1,\lsd]$.

@@

For the first, if we constrain the scale matrix to be proportional to the identity matrix, $\Ppscale = \Ppscalescalar \idmat$, then it can be shown (see \cite{Uhlig1994,Srivastava2003} and \cite[chapter 3]{Muirhead1982}) that the eigenvectors and eigenvalues of $\lgtp$ are independent. Furthermore, the distribution for the eigenvector matrix is uniform (i.e. a normalised Haar measure, conditional on the first element being positive), and for the eigenvalues,
%
\begin{IEEEeqnarray}{rCl}
 \den(\tnmd|\tnd) & = & \frac{ \pi^{\half\tnd^2} }{ (2\Ppscalescalar)^{\half\tnd\lsd} \gammafun[\tnd](\frac{\tnd}{2}) \gammafun[\tnd](\frac{\lsd}{2}) } \left[ \prod_{i<j}^{\tnd} (\tnmev{i}-\tnmev{j}) \right] \nonumber \\
 & & \qquad \times \left[ \prod_{i=1}^{\tnd} \tnmev{i}^{\half(\lsd-\tnd-1)} \exp\left(-\frac{1}{2\Ppscalescalar} \tnmev{i}\right) \right]      . \nonumber \\ \label{eq:eigenvalue_prior}
\end{IEEEeqnarray}

For the second factorisation of $\lgtp$, since $\tnmf = \tnmocrossred\tr \lgtp \tnmocrossred$, we find that,
%
\begin{IEEEeqnarray}{rCl}
 \den(\tnmf|\tnmocrossred, \tnd) & = & \wishartden{\tnmf}{\tnd}{\Ppscalescalar \idmat}      .
\end{IEEEeqnarray}
%
(See \cite[theorem 3.2.5]{Muirhead1982}.) It is possible to derive a corresponding prior distribution for a parameterisation of $\tnmocrossred$ using a set of Givens rotations, but this will not be needed.

\subsection{MCMC for Degenerate Transitions}

As discussed above, standard Gibbs sampling from $\postden(\lgtp|\lgtm,\ls{1:\timax})$ and $\postden(\lgtm|\lgtp,\ls{1:\timax})$ does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\tnd$-dimensional column space of $\lgdm$. Nevertheless, we can sample from these distributions exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions.

In addition to the within-subspace Gibbs moves, we introduce three efficient Metropolis-Hastings moves which target $\postden(\lgtp|\lgtm)$ and $\postden(\lgtm, \ls{1:\timax}|\lgtp)$, thus allowing the transition subspace to be changed and the parameter space to be fully explored by the Markov chain.

\subsubsection{Gibbs Sampling within the Subspace}

It is here that we exploit the transformation from $\lgtp$ to $\{\tnmocrossred,\tnmf\}$ introduced in \eqref{eq:degenerate_covariance_factorisation}. In the degenerate case, the transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtp)$ with respect to the Lebesgue measure does not exist. However, since the set of possible values for $\ls{\ti}$ constitutes a hyperplane, we can project this onto a lower dimensional space in which a density does exist. From \eqref{eq:degenerate_transition} and \eqref{eq:degenerate_covariance_factorisation}, 
%
\begin{IEEEeqnarray}{rCl}
 \tnmocrossred\tr(\ls{\ti}-\lgtm\ls{\ti-1}) & \sim & \normaldist{0}{\tnmf\inv}        .
\end{IEEEeqnarray}
%
The likelihood of the state trajectory can then be written as,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm,\tnmf,\tnmocross) } \nonumber \\
 & \propto & \begin{cases}
                                  \determ{\tnmf}^{\frac{\timax-1}{2}} \exp\left( -\half \sum\limits_{\ti=1}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr\tnmocrossred \tnmf \tnmocrossred^T(\ls{\ti}-\lgtm\ls{\ti-1}) \right) \\ \hphantom{0} \qquad \text{if }\forall \,\, \ti, (\ls{\ti}-\lgtm\ls{\ti-1}) \text{ lies in the column space of } \tnmocrossred \\
                                  0 \qquad \text{otherwise}     .
                                                        \end{cases} \nonumber 
\end{IEEEeqnarray}

We now re-formulate the posterior conditional for $\lgtp$ in terms of $\tnmocrossred$ and $\tnmf$,
%
\begin{IEEEeqnarray}{rCl}
 \postden(\tnmocrossred, \tnmf | \lgtm,\ls{1:\timax}) & = & \postden(\tnmf | \tnmocrossred, \lgtm,\ls{1:\timax}) \postden(\tnmocrossred | \lgtm,\ls{1:\timax})    .
\end{IEEEeqnarray}
%
Because of the subspace constraint, $\tnmocrossred$ is uniquely defined conditional on $\ls{1:\timax}$, and thus sampling from $\postden(\tnmocrossred | \lgtm,\ls{1:\timax})$ simply implies keeping the existing value of $\tnmocrossred$. This can be proved using a ``degrees of freedom'' argument. The set of points $\{(\ls{\ti}-\lgtm\ls{\ti-1})\}_{t=1}^{\timax}$ lie in a $\tnd$-dimensional subspace, characterised by $\tnd(\lsd-\tnd)$ independent parameters. This is the same number as the parameters used to describe $\tnmocrossred$. (See appendix~\ref{app:givens-factorisation}.)

For $\tnmf$ the appropriate conditional is,
%
\begin{IEEEeqnarray}{rCl}
 \postden(\tnmf|\lgtm, \tnmocrossred, \tnd, \ls{1:\timax}) &\propto& \den(\ls{1:\timax}|\lgtm,\tnmf,\tnmocrossred) \den(\tnmf) \nonumber \\
% \qquad & \propto & \determ{\tnmf}^{\frac{\timax-2}{2}} \exp\left( -\half \trace\left[ \tnmf \left( \frac{1}{\Ppscalescalar}\idmat + \sum_{\ti=2}^{\timax} \tnmocrossred\tr(\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr\tnmocrossred \right) \right] \right)  \nonumber \\
 & = & \wishartden{\tnmf}{\Ppdof''}{\Ppscale''}      ,
\end{IEEEeqnarray}
%
where,
%
\begin{IEEEeqnarray}{rCl}
 \Ppdof''   & = & \tnd + \timax - 1 \nonumber \\
 \Ppscale'' & = & \left[\frac{1}{\Ppscalescalar}\idmat + \sum_{\ti=2}^{\timax} \tnmocrossred\tr(\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr\tnmocrossred \right]\inv \label{eq:D_posterior_params}      ,
\end{IEEEeqnarray}
%
which closely resembles that for the precision matrix in the full rank case.

>>>>

The Gibbs sampler approach developed in \cite{Wills2012} is effective on the basic flavour of linear Gaussian state space models. However, it is not able to handle the case in which the transition model is degenerate, i.e. when the transition covariance matrix is not full rank. In this paper we extend the MCMC learning framework to encompass this scenario. This is achieved by appropriate factorisations of the covariance matrix which allow us still to exploit the fast-mixing Gibbs kernel. Furthermore, reversible jump construction \cite{Green1995,Green2009} is introduced which enables the sampler to learn the rank of the transition covariance matrix.

@@

\subsection{Gibbs Sampling for Basic Linear Gaussian Models}

@@

\section{Degenerate Transition Models} \label{sec:degenerate_transition_models}

A commonly encountered variation on the basic linear Gaussian state space model is as follows,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \lgdm \tn{\ti} \label{eq:degenerate_transition} \\
 \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\idmat}      ,
\end{IEEEeqnarray}
%
in which $\tn{\ti}$ as dimension $\rk$, and $\lgdm$ is $\lsd\times\rk$. This is almost equivalent to the previous, with $\lgtv=\lgdm\lgdm\tr$. However, whereas before $\lgtv$ was required to be positive definite, this is no longer the case. When this occurs, the conventional transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgdm)$ is not defined, and the sampling operations underlying our Gibbs sampler are no longer valid. Degenerate models such as these arise in tracking applications (see \cite{Maskell2004} for a discussion of this phenomenon); when non-Markovian models, such as autoregressive process, are converted to state space form; and when it is natural to parameterise a system with more latent states than there are degrees of freedom (see section~\ref{sec:mocap}). In this section, we introduce a parameterisation for such degenerate transition models which can uniquely represent any such system, and show how MCMC moves may be implemented for learning.

The matrix $\lgdm$ is not uniquely identifiable from the state sequence, because the distribution of $\lgdm\tn{\ti}$ is identical to that of $\lgdm \Xi \tn{\ti}$ where $\Xi$ is an arbitrary orthogonal matrix. To avoid this ambiguity, it is better to focus on learning $\lgtv$ as before, and to define $\lgdm$ as some matrix square root of $\lgtv$. In this new setting, $\lgtv$ is a positive semi-definite matrix with rank $\rk$.

There are two difficulties with extending the methodology from the basic model to the degenerate case. First, the conjugate matrix normal--inverse Wishart distribution becomes singular. This places an undesirable constraint on the transition matrix which needs to be removed. Second, each transition, $(\ls{\ti}-\lgtm\ls{\ti-1})$, is now constrained to lie in a particular $\rk$-dimensional subspace within the state space; specifically, within the column space of $\lgdm$. This subspace is fixed conditional on any sampled state trajectory $\ls{1:\timax}$ (provided that $\timax>\rk$). This makes standard Gibbs sampling from $\postden(\lgtm, \lgtv|\ls{1:\timax})$ insufficient, since there are values of $\lgtm$ and $\lgtv$ which can never be reached.

There is an obvious way to avoid being constrained to a particular subspace by the sampled state trajectory; simply return to targeting the marginal posterior $\postden(\lgtm, \lgtv|\ob{1:\timax})$ and fall back to using a generic Metropolis-Hastings approach. However, with a complex and high dimensional parameter space, designing efficient proposal distributions for Metropolis-Hastings is difficult. We would prefer to attain the automated, fast-mixing properties displayed by the Gibbs sampler on the full rank model. In this section we show how a Gibbs kernel may still be applied to sample a component of the parameters, with the remainder handled using Metropolis-Hastings.

In formulating an MCMC algorithm for degenerate transition models, we will make use of two factorisations of the transition covariance matrix. First, an eigendecomposition,
%
\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \tvvec \tvval \tvvec\tr     ,
\end{IEEEeqnarray}
%
in which $\tvval$ is a diagonal $\rk\times\rk$ matrix with $\rk = \rank(\lgtv)$, and $\tvvec$ is a $\lsd\times\rk$ matrix of orthonormal columns from the appropriate Stiefel manifold. This is the ``non-singular'' part of the factorisation. In order to resolve ambiguities in this representation, we follow \cite{Muirhead1982} and require that the first element in each column of $\tvval$ should be positive, and that eigenvalues in $\tvval$ be arranged in decreasing order of magnitude. The transformation from $\lgtv$ to $\{\tvval, \tvvec\}$ is then bijective.

We can also write the ``full'' factorisation by finding a set of $\lsd-\rk$ additional orthonormal vectors and appending them to $\tvvec$, along with matching zeros to $\tvval$, as follows,
%
\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \begin{bmatrix}\tvvec & \tvvecorth \end{bmatrix} \begin{bmatrix}\tvval & \zmat \\ \zmat & \zmat \end{bmatrix} \begin{bmatrix}\tvvec\tr \\ \tvvecorth\tr \end{bmatrix}
\end{IEEEeqnarray}
%
Note that $\tvvecorth$ is not unique.

A further factorisation will also be useful, which we will refer to as the Givens decomposition (on account of its method of computation). For the rank-$\rk$ positive semi-definite covariance matrix $\lgtp$,
%
\begin{IEEEeqnarray}{rClCl}
 \lgtv &=& \tvrot \tvfull \tvrot\tr &=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix} \label{eq:givens_decomposition}
\end{IEEEeqnarray}
%
where,
%
$\tvfull$ is a $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors subject to certain constraints. The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective. See appendix~\ref{app:givens-factorisation} for details.

@@

The matrix normal--inverse Wishart conjugate prior will need some modifications in order to apply it to degenerate models.

An appropriate prior distribution for the degenerate transition covariance matrix may be defined by setting the degrees of freedom of an inverse Wishart distribution to the desired rank, $\priordof=\rk$. This results in a singular inverse Wishart distribution over the rank-$\rk$ positive semi-definite matrices \cite{Diaz-Garcia2006}. The density (with respect to the Lesbegue measure on the distinct elements),
%
\begin{IEEEeqnarray}{rCl}
 \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrix}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrix \right)    ,
\end{IEEEeqnarray}
%
in which $\lgtv\pinv$ denotes the Moore-Penrose pseudoinverse of $\lgtv$.

The matrix normal component of the prior is more problematic. Although the distribution it is still well-defined, the fact that $\lgtv$ is singular means that the space of possible values is restricted to a subspace of $\reals^{\lsd\times\lsd}$, placing an unwanted constraint on our model. (See \cite{Diaz-Garcia2006} for a discussion of the singular matrix normal distribution.)

In order to relax this unwanted constraint, the following non-singular prior distribution may be used,
%
\begin{IEEEeqnarray}{rCl}
 \lgtm|\lgtv &\sim& \matrixnormaldist{\priormeanmatrix}{\lgtv+\tvvecorth\priorextratpval\tvvecorth\tr}{\priorcolumnvariance}
\end{IEEEeqnarray}
%
in which $\priorextratpval$ is a diagonal matrix of positive eigenvalues. The matrix $\lgtv+\tvvecorth\priorextratpval\tvvecorth\tr$ is then guaranteed to be positive definite.

Using this prior, we need a mechanism for choosing $\priorextratpval$. These extra eigenvalues control the rate at which probability decays as the matrix moves away from the subspace defined by $\lgtv$.
\meta{Add something about how to set these.}

Finally, the prior is completed by specifying a distribution over the possible values of the rank $\den(\rk)$.

\subsection{Constrained Gibbs Sampling for Degenerate Transitions}

As discussed, standard Gibbs sampling using $\postden(\lgtm, \lgtv|\ls{1:\timax})$ does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\rk$-dimensional column space of $\lgdm$. Nevertheless, we can sample from these distributions exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions. These constrained Gibbs sampling steps rely on the Givens decomposition \eqref{eq:givens_decomposition}.

\subsubsection{Likelihood and Subspace Constraints}
There is no transition density over $\reals^{\lsd}$ for the degenerate model. However, a density does exist with respect to an appropriate measure on the reachable subspace, which provides us with a likelihood function \cite{Diaz-Garcia2006}. We can derive this from the underlying state transition equation,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} &=& \lgtm \ls{\ti-1} + \tvrot \tvfull\msqrt \tn{\ti} \nonumber
 \end{IEEEeqnarray}
 \begin{IEEEeqnarray}{rCl}
 \Rightarrow \tvrot\tr(\ls{\ti}-\lgtm\ls{\ti-1}) &\sim& \normaldist{\zmat}{\tvfull} \nonumber
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) &\propto& \determ{\tvfull}^{-\half} \exp\left(-\half (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right)     .
\end{IEEEeqnarray}
%
The subspace over which this is defined is,
%
\begin{IEEEeqnarray}{rCl}
 \{\ls{\ti} : \ls{\ti} = \lgtm \ls{\ti-1} + \tvrot z, z \in \reals^{\rk}\}     .
\end{IEEEeqnarray}

The likelihood function for the entire state trajectory is thus,
%
\begin{align}
 \den(\ls{1:\timax}|\lgtm, \lgtv) &= \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) \nonumber \\
 &\propto \determ{\tvfull}^{-\half(\timax-1)} \exp\left( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber \\
 &= \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
 &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
 &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right)
\end{align}
%
in which the sufficient statistics are as before and,
%
\begin{equation}
 \lgtmrot = \tvrot\tr \lgtm      .
\end{equation}

Writing all the subspace constraints concisely in a single equation,
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tvrot Z
\end{IEEEeqnarray}
%
where $Z \in \reals^{\rk\times\lsd}$ and 
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \begin{bmatrix} \ls{2} & \ls{3} & \hdots & \ls{\timax} \end{bmatrix}    .
\end{IEEEeqnarray}

Now introduce,
%
\begin{IEEEeqnarray}{rCl}
 \lgtmorth &=& (\idmat-\tvrot\tvrot\tr) \lgtm     ,
\end{IEEEeqnarray}
%
such that the transition matrix may be decomposed into orthogonal components,
%
\begin{IEEEeqnarray}{rCl}
 \lgtm &=& \tvrot \lgtmrot + \lgtmorth    .
\end{IEEEeqnarray}
%
Using this, the constraint equation becomes,
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& (\tvrot \lgtmrot + \lgtmorth) \LS{1:\timax-1} + \tvrot Z \nonumber \\
 \Rightarrow \LS{2:\timax} &=& \lgtmorth \LS{1:\timax-1} + \tvrot Z' \nonumber      .
\end{IEEEeqnarray}
%
Notice that $\lgtmrot$ and $\tvfull$ do not appear in this equation, and thus may be freely altered by the sampler. In contrast, $\lgtmorth$ and $\tvrot$ are uniquely determined. Sampling from $\postden(\lgtm,\lgtv|\ls{1:\timax})$ thus entails drawing new values from $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$.

\subsubsection{Transforming the Prior}
We will need the distribution on $\{\lgtmrot, \tvfull\}$ implied by the prior we chose for $\{\lgtm,\lgtv\}$. These follow from simple properties of the wishart and matrix normal distributions (see \cite{Muirhead1982}). For the matrix normal part,
%
\begin{IEEEeqnarray}{rCl}
 \lgtmrot|\tvrot,\tvfull &\sim& \matrixnormaldist{\tvrot\tr\priormeanmatrix}{\tvfull}{\priorcolumnvariance}     .
\end{IEEEeqnarray}
%
For the inverse Wishart part,
%
\begin{IEEEeqnarray}{rCl}
 \tvfull|\tvrot &\sim& \iwishartdist{\rk}{(\tvrot\tr\priorscalematrix\inv\tvrot)\inv}     .
\end{IEEEeqnarray}
%
Notice that this is no longer a singular Inverse Wishart distribution, since $\tvfull$ is $\rk\times\rk$.

\subsubsection{Sampling the Conditional}
Using the constrained likelihood function and the transformed prior, we now obtain the required posterior conditional distribution $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$. This follows the same algebraic steps as in the full rank case and leads to a similar matrix normal--inverse Wishart distribution,
%
\begin{align}
 \tvfull | \tvrot, \ls{1:\timax} &\sim \iwishartdist{\postdof}{\postscalematrix} \\
 \lgtmrot | \tvfull, \tvrot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}     ,
\end{align}
%
with the following updated hyperparameters,
%
\begin{align}
 \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
 \postmeanmatrix \postcolumnvariance\inv &= \tvrot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
 \postdof                                &= \rk + \suffstats{0} \\
 \postscalematrix                        &= (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\left( \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tvrot - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
\end{align}

Once new values of $\lgtmrot$ and $\tvfull$ have been sampled, the corresponding values of $\lgtm$ and $\lgtv$ may be calculated using the existing $\lgtmorth$ and $\tvrot$.

\subsection{Metropolis-Hastings on the Posterior Marginal}

Gibbs sampling is very efficient and requires no algorithm tuning but it only allows the Markov chain to explore a subset of the parameter space. In this section we introduce two Metropolis-Hastings kernels which target the marginal posterior $\postden(\lgtm,\lgtv)$ and allow the sampler to fully explore the target distribution. These sampling operations are each followed by a draw of the state trajectory from $\postden(\ls{1:\timax}|\lgtm,\lgtv)$, thus constituting a collapsed Gibbs move \cite{Dyk2008}.

\subsubsection{Covariance Random Walk}

The covariance matrix is specified by its eigenvalues and eigenvectors. The sampler is already able to explore the possible eigenvalues using the constrained Gibbs moves, but the eigenvector matrix is partially fixed during this process. We can allow the sampler to reach any possible eigenvector matrix using a rotational random walk. Suppose the current value of the transition covariance is $\lgtv\mcold$, a new value of $\lgtv$ is proposed using the following procedure:
%
\begin{itemize}
 \item Sample an orthogonal matrix $\mhrot$ from some distribution $\mhrotppsl$.
 \item Set $\lgtv\mcnew = \mhrot \lgtv\mcold \mhrot\tr$.
\end{itemize}
%
The transformation is invertible and has a Jacobian of $1$. If in addition we require that $\mhrotppsl(\mhrot) = \mhrotppsl(\mhrot\tr)$, then the proposal distribution is symmetric and the acceptance probability is simply,
%
\begin{IEEEeqnarray}{rCl}
 \mhap(\lgtp\mcold\to\lgtp\mcnew) & = & \min\left\{1, \frac{ \postden(\lgtm, \lgtp\mcnew) }{ \postden(\lgtm,\lgtp\mcold) } \right\} \\
  & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtp\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtp\mcold)} \times \frac{\den(\lgtm,\lgtp\mcnew)}{\den(\lgtm,\lgtp\mcold)} \right\} \nonumber     .
\end{IEEEeqnarray}

There are numerous ways to sample the rotation matrix $\mhrot$ from a suitable proposal distribution $\mhrotppsl$. For example, we could use the Cayley transform \cite{Leon2006}, a bijective mapping from the skew-symmetric matrices to the rotation matrices, defined by,
%
\begin{IEEEeqnarray}{rCl}
 P(S) & = & (\idmat - S)\inv(I+S)     .
\end{IEEEeqnarray}
%
To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a nice choice is,
%
\begin{IEEEeqnarray}{rCl}
 s_{k,l} & \sim & \normaldist{0}{\sigma_s^2} \label{eq:skewsymmetric_proposal}     .
\end{IEEEeqnarray}
%
Use these to construct a skew-symmetric matrix $S$,
%
\begin{IEEEeqnarray}{rCl}
 S_{k,l} & = & \begin{cases}
                s_{k,l}  & k<l \\
                -s_{l,k} & k>l \\
                0        & k=l     ,
               \end{cases}
\end{IEEEeqnarray}
%
and then set $\mhrot=P(S)$. The Cayley transform has the property that $P(-S)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)$ as required.

There is an alternative using Givens rotations. First sample $i \in [1,\lsd]$, $j \in [1,\lsd]\setminus i$, and $\givrot{} \in [-\pi/2,\pi/2]$ from some zero-mean distribution. Form the Givens matrix $\givmat{i}{j}{\givrot{}}$ such that,
%
\begin{IEEEeqnarray}{rCl}
 \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
                                                    \sin(\givrot{}) & k=i,l=j \\
                                                    -\sin(\givrot{}) & k=j,l=i \\
                                                    0 & \text{ otherwise,}
                                                 \end{cases}
\end{IEEEeqnarray}
%
and use $\mhrot=\givmat{i}{j}{\givrot{}}$. This also has the property that $\givmat{i}{j}{-\givrot{}} = \givmat{i}{j}{\givrot{}}\tr$, implying a symmetric proposal.

\subsubsection{Transition Matrix Random Walk}

\subsection{Reversible Jump for Rank Learning}

\section{Simulations}

\subsection{A Toy Model}

\subsection{Motion Capture Interpolation}

%%% REWRITING %%%

\section{Old Material}

\subsubsection{Transforming the Prior}

@@

For the first, if we constrain the scale matrix to be proportional to the identity matrix, $\Ppscale = \Ppscalescalar \idmat$, then it can be shown (see \cite{Uhlig1994,Srivastava2003} and \cite[chapter 3]{Muirhead1982}) that the eigenvectors and eigenvalues of $\lgtp$ are independent. Furthermore, the distribution for the eigenvector matrix is uniform (i.e. a normalised Haar measure, conditional on the first element being positive), and for the eigenvalues,
%
\begin{IEEEeqnarray}{rCl}
 \den(\tnmd|\rk) & = & \frac{ \pi^{\half\rk^2} }{ (2\Ppscalescalar)^{\half\rk\lsd} \gammafun[\rk](\frac{\rk}{2}) \gammafun[\rk](\frac{\lsd}{2}) } \left[ \prod_{i<j}^{\rk} (\tnmev{i}-\tnmev{j}) \right] \nonumber \\
 & & \qquad \times \left[ \prod_{i=1}^{\rk} \tnmev{i}^{\half(\lsd-\rk-1)} \exp\left(-\frac{1}{2\Ppscalescalar} \tnmev{i}\right) \right]      . \nonumber \\ \label{eq:eigenvalue_prior}
\end{IEEEeqnarray}
